Model Search in Relational Domains

### 2016-10-23 domains.tex
wording

%-------------------------------------------------------------------------------
\subsection{Program Representation}  
\Omit {
\begin{wrapfigure}{r}{3cm}
\[
\begin{array}{rcl}
g_0 &=& \true \\
\multicolumn{3}{l}{0 \leq v\leq N} \\
g_1 &=& g_0\wedge c\\
x_0 &=& v \\
x_1 &=& -v \\
x_2 &=& g_1?x_0:x_1 \\
g_2 &=& g_1 \vee g_0\wedge \neg c\\
z &=& x_2 * x_2 \\
\multicolumn{3}{l}{z<0 \wedge g_2}
\end{array}
\]
\caption{\label{fig:ssa}
SSA for the example in Figure~\ref{fig:example}
}
\end{wrapfigure}
}
\begin{figure}[t]
\center
\scriptsize
\begin{tabular}{l|l}
\hline
 SSA & Abstract Conflict Graph \\
\hline
\begin{minipage}{4.50cm}
$\begin{array}{rcl}
\constraints:= (g_0 = (0 \leq v\leq N)) \\
(g_1 = (g_0 \wedge c)) \wedge (x_0 = v) \wedge \\
(x_1 = -v) \wedge (x_2 = g_1?x_0:x_1) \wedge \\
(g_2 = (g_1 \vee g_0\wedge \neg c)) \wedge \\
(z  = x_2 * x_2) \wedge (g_2 \wedge z<0)
\end{array}$
\end{minipage}
&
\begin{minipage}{5.2cm}
\centering
\scalebox{.45}{\import{figures/}{learning.pspdftex}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{SSA and abstract conflict graph}
\label{ssa}
\end{figure}
%
In this paper, we consider \emph{bounded programs} with safety
properties given as assertions $\assertions$ in the program.
%
A bounded program is obtained by a program transformation that unfolds
loops and recursions a finite number of times. The result of this
transformation is represented as set
$\constraints=\program\cup\{\neg \bigwedge_{\assertion\in\assertions} \assertion\}$
where $\program$ are the statements of the transformed program in the
form of constraints as typically obtained by translating the program
into single static assignment (SSA) form via a data flow analysis.
%
Figure~\ref{ssa} shows the SSA constraints for the program in
Figure~\ref{fig:example}.
%
Assignments such as \texttt{x=v;} become equalities $x_1=v$ where the
left-hand side variable gets a fresh name (subscript).
%
Control flow is encoded into guard variables such as $g_1=g_0\wedge c$.
%
Data flow joins become conditional expressions, e.g.\ $x_3=g_1?x_1:x_2$.
%
The assertions $\assertions$ are constraints of the form $g_3
\Rightarrow z\geq 0$, for instance, meaning that if $g_3$ holds
(i.e.\ the assertion is reachable) then the assertion must hold.
%
We denote $\vars$ the set of variables occurring in the $\constraints$.
%
%
%For example, figure~\ref{swssa} presents the program analysis constraints for 
%a C program in Static Single Assignment (SSA) form.

@@

% \paragraph{Interval Abstract Domain} 
% The {\em Interval} abstract domain is a lattice 
% $ItvDom = (ItvElm, \sle_{I}, \join_{I}, \meet_{I})$, where
% $ItvElm: (Var \rightarrow Itv) \cup \{\bot\}$, and $Itv$ is 
% the set of intervals of type $[l,u]$ over numeric data 
% type with $l \leq u$. The least element is $\bot$ and the 
% greatest element is $\top$ which maps all variables to their
% minimum ($min$) and maximum ($max$) values.  An interval 
% $\langle x, [min, v] \rangle$ is written as $x \leq v$.  The 
% partial order $\sle_{I}$ over elements in the set $Itv$ is 
% given by $I_1 \sle_{I} I_2$ if $I_2$ contains $I_1$.
% A join $(\join_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [min(l_1, l_2), max(u_1, u_2)]$.
% A meet $(\meet_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [max(l_1, l_2), min(u_1, u_2)] \rangle$.
% The galois connection between the concrete environment and intervals is 
% given as follows.
% \[\alpha(x) = \{[inf(x), sup(x)] | x \in Var\} \qquad   \alpha(\emptyset) = \bot \]
% \[\gamma[p,q] = \{c \in \mathbb{Z} | p \leq c \leq q\} \qquad \gamma(\bot) = \emptyset \]
%
\begin{wrapfigure}{r}{4.5cm}
\vspace*{-7ex}
\scalebox{.65}{\import{figures/}{octagons.pspdftex}}
\caption{Example of an octagon}\label{octagon}
\vspace*{-5ex}
\end{wrapfigure} 
%
In the interval domain $\intervals[\numvars]$,
for a program with $N=|\numvars|$ variables, the matrix $\mat{C}$ 
has $2N$ rows; and hence it generates at most $2N$ inequalities, one
for the upper and lower bounds of each variable.
%
For octagons $\octagons[\numvars]$, we have at most $2N^2$
inequalities, one for the upper and lower bounds of each variable and
sums and differences for each pair of variables. Figure~\ref{octagon}
shows an example octagon and its associated inequalities.

@@

%-------------------------------------------------------------------------------
\paragraph{\textbf{Abstract Transformers}}
%
%In a typical abstract interpretation based Galois-connection setting
%over an over-approximate domain, every concrete element has a unique
%over-approximate representation in the abstract.  Likewise, every
%concrete transformer is over-approximated by a unique abstract
%transformer.  We now define an abstract deduction transformer.
%
An abstract transformer $\abstrans{\domain}{\constraint}$ transforms an abstract value $\absval$ through a
constraint $\constraint$, i.e.\ it \emph{deduces} information from $\absval$ and $\constraint$.
The best abstract transformer is 
\begin{equation}\label{eq:abstrans}
\abstrans{\domain}{\constraint}(\absval)=\absval\meet\alpha(\{\val\mid \val\in\gamma(\absval), \val\models \constraint\})
\end{equation}
 where we write 
$\val\models \constraint$ if the concrete value $\val$ satisfies the constraint $\constraint$.
%$\forall \vec{y}\in c: (s[\vec{y}/\vec{\numvar}]
%\equiv \true)$, i.e.\ the constraint $s$ is valid when evaluated over
%all values $\vec{y}$ of its variables $\vec{\numvar}$ in the concretisation
%of $a$. 
Any abstract transformer that over-approximates the best abstract
transformer is a sound transformer and can be used in our algorithm.
%
For example, we can deduce $\abstrans{\domain}{x=2(y+z)}(\absval)=\absval\meet(-2\leq x\leq 6)$ for the abstract value $\absval=(0\leq y \leq 2 \wedge 1 \leq y-z \leq 1)$.
%
%A detailed description of abstract transformer is presented in Section~\ref{sec:abst}.
We denote the set of abstract transformers for a safety equation
$\formula$ using abstract domain $\domain$ as
$\abstransset=\{\abstrans{\domain}{\constraint}\mid
\constraint\in\constraints\}$.

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
An important property of a clause learning SAT solver is that each
non-singleton element of the partial assignment domain can be 
decomposed into a set of precisely complementable singleton elements.  
This property of the partial assignment domain is required to learn 
elements that help to guide the model search away from the
conflicting region of the search space.  Complementation operator 
in abstract domains is different from the notion of precise complements. 
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
half-spaces, each of which has a complement. 
%
% We identify specific properties of domain elements necessary for 
% abstract model search and learning in abstractions.  
We formalise this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\begin{equation}
\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top  
\end{equation}
Meet irreducibles in ACDCL correspond to the concept of literals in a SAT solver.
\end{definition}
%
The meet irreducibles in the Boolean domain $\booldomain$ for a variable $x$ are 
$x$ and $\neg x$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e.\ they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.

@@

\begin{definition} 
An element $\absval\in \domain$ is called \emph{precisely complementable}
iff $\exists \bar{\absval} \in \domain: \neg\gamma(\bar{\absval})=\gamma(\absval)$, 
i.e.\ there is an element whose negated concretisation equals
the concretisation of $\absval$.
%m\meet\bar{m}=\bot \wedge m\join\bar{m}=\top$.
%A complementable meet irreducible $\bar{M}$ of an abstract lattice $A$ is the complement of a meet 
%irreducible $M \in A$ such that $\bar{M} \in A$ and the concretisation of $M$ 
%is the set complement of $\bar{M}$.  
%If every meet irreducible in $A$ is complementable, then $A$ is said to
%be have complementable meet irreducibles.
\end{definition}

>>>>

%-------------------------------------------------------------------------------
\subsection{Program Representation}  
\Omit {
\begin{wrapfigure}{r}{3cm}
\[
\begin{array}{rcl}
g_0 &=& \true \\
\multicolumn{3}{l}{0 \leq v\leq N} \\
g_1 &=& g_0\wedge c\\
x_0 &=& v \\
x_1 &=& -v \\
x_2 &=& g_1?x_0:x_1 \\
g_2 &=& g_1 \vee g_0\wedge \neg c\\
z &=& x_2 * x_2 \\
\multicolumn{3}{l}{z<0 \wedge g_2}
\end{array}
\]
\caption{\label{fig:ssa}
SSA for the example in Figure~\ref{fig:example}
}
\end{wrapfigure}
}
\begin{figure}[t]
\center
\scriptsize
\begin{tabular}{l|l}
\hline
 SSA & Abstract Conflict Graph \\
\hline
\begin{minipage}{4.50cm}
$\begin{array}{rcl}
\constraints:= (g_0 = (0 \leq v\leq N)) \\
(g_1 = (g_0 \wedge c)) \wedge (x_0 = v) \wedge \\
(x_1 = -v) \wedge (x_2 = g_1?x_0:x_1) \wedge \\
(g_2 = (g_1 \vee g_0\wedge \neg c)) \wedge \\
(z  = x_2 * x_2) \wedge (g_2 \wedge z<0)
\end{array}$
\end{minipage}
&
\begin{minipage}{5.2cm}
\centering
\scalebox{.45}{\import{figures/}{learning.pspdftex}}
\end{minipage}
\\
\hline
\end{tabular}
\caption{SSA and abstract conflict graph}
\label{ssa}
\end{figure}
%
In this paper, we consider \emph{bounded programs} with safety
properties given as assertions $\assertions$ in the program.
%
A bounded program is obtained by a program transformation that unfolds
loops and recursions a finite number of times. The result of this
transformation is represented as set
$\constraints=\program\cup\{\neg \bigwedge_{\assertion\in\assertions} \assertion\}$
where $\program$ are the statements of the transformed program in the
form of constraints as typically obtained by translating the program
into single static assignment (SSA) form via a data flow analysis.
%
Figure~\ref{ssa} gives the SSA constraints for the program in
Figure~\ref{fig:example}.
%
Assignments such as \texttt{x=v;} become equalities $x_1=v$ where the
left-hand side variable gets a fresh name (subscript).
%
Control flow is encoded into guard variables such as $g_1=g_0\wedge c$.
%
Data flow joins become conditional expressions, e.g.\ $x_3=g_1?x_1:x_2$.
%
The assertions $\assertions$ are constraints of the form $g_3
\Rightarrow z\geq 0$, for instance, meaning that if $g_3$ holds
(i.e.\ the assertion is reachable) then the assertion must hold.
%
We denote $\vars$ the set of variables occurring in the $\constraints$.
%
%
%For example, figure~\ref{swssa} presents the program analysis constraints for 
%a C program in Static Single Assignment (SSA) form.

@@

% \paragraph{Interval Abstract Domain} 
% The {\em Interval} abstract domain is a lattice 
% $ItvDom = (ItvElm, \sle_{I}, \join_{I}, \meet_{I})$, where
% $ItvElm: (Var \rightarrow Itv) \cup \{\bot\}$, and $Itv$ is 
% the set of intervals of type $[l,u]$ over numeric data 
% type with $l \leq u$. The least element is $\bot$ and the 
% greatest element is $\top$ which maps all variables to their
% minimum ($min$) and maximum ($max$) values.  An interval 
% $\langle x, [min, v] \rangle$ is written as $x \leq v$.  The 
% partial order $\sle_{I}$ over elements in the set $Itv$ is 
% given by $I_1 \sle_{I} I_2$ if $I_2$ contains $I_1$.
% A join $(\join_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [min(l_1, l_2), max(u_1, u_2)]$.
% A meet $(\meet_{I})$ of two intervals $\langle x_1 \rightarrow [l_1, u_1], 
% x_1 \rightarrow [l_2, u_2] \rangle$ is an interval 
% $\langle x_1 \rightarrow [max(l_1, l_2), min(u_1, u_2)] \rangle$.
% The galois connection between the concrete environment and intervals is 
% given as follows.
% \[\alpha(x) = \{[inf(x), sup(x)] | x \in Var\} \qquad   \alpha(\emptyset) = \bot \]
% \[\gamma[p,q] = \{c \in \mathbb{Z} | p \leq c \leq q\} \qquad \gamma(\bot) = \emptyset \]
%
\begin{wrapfigure}{r}{4.5cm}
\vspace*{-7ex}
\scalebox{.65}{\import{figures/}{octagons.pspdftex}}
\caption{Example of an octagon}\label{octagon}
\vspace*{-5ex}
\end{wrapfigure} 
%
In the interval domain $\intervals[\numvars]$,
for a program with $N=|\numvars|$ variables, the matrix $\mat{C}$ 
has $2N$ rows; and hence it generates at most $2N$ inequalities, one
for the upper and lower bounds of each variable.
%
For octagons $\octagons[\numvars]$, we have at most $2N^2$
inequalities, one for the upper and lower bounds of each variable and
sums and differences for each pair of variables. Figure~\ref{octagon}
gives an example of an octagon and its associated inequalities.

@@

%-------------------------------------------------------------------------------
\paragraph{\textbf{Abstract Transformers}}
%
%In a typical abstract interpretation based Galois-connection setting
%over an over-approximate domain, every concrete element has a unique
%over-approximate representation in the abstract.  Likewise, every
%concrete transformer is over-approximated by a unique abstract
%transformer.  We now define an abstract deduction transformer.
%
An abstract transformer $\abstrans{\domain}{\constraint}$ transforms an abstract value $\absval$ through a
constraint $\constraint$, i.e., it \emph{deduces} information from $\absval$ and $\constraint$.
The best abstract transformer is 
\begin{equation}\label{eq:abstrans}
\abstrans{\domain}{\constraint}(\absval)=\absval\meet\alpha(\{\val\mid \val\in\gamma(\absval), \val\models \constraint\})
\end{equation}
 where we write 
$\val\models \constraint$ if the concrete value $\val$ satisfies the constraint $\constraint$.
%$\forall \vec{y}\in c: (s[\vec{y}/\vec{\numvar}]
%\equiv \true)$, i.e., the constraint $s$ is valid when evaluated over
%all values $\vec{y}$ of its variables $\vec{\numvar}$ in the concretisation
%of $a$. 
Any abstract transformer that over-approximates the best abstract
transformer is a sound transformer and can be used in our algorithm.
%
For example, we can deduce $\abstrans{\domain}{x=2(y+z)}(\absval)=\absval\meet(-2\leq x\leq 6)$ for the abstract value $\absval=(0\leq y \leq 2 \wedge 1 \leq y-z \leq 1)$.
%
%A detailed description of abstract transformer is presented in Section~\ref{sec:abst}.
We denote the set of abstract transformers for a safety equation
$\formula$ using abstract domain $\domain$ as
$\abstransset=\{\abstrans{\domain}{\constraint}\mid
\constraint\in\constraints\}$.

%-------------------------------------------------------------------------------
\subsection{Properties of Abstract Domains}
%-------------------------------------------------------------------------------
%
An important property of a clause learning SAT solver is that each
non-singleton element of the partial assignment domain can be 
decomposed into a set of precisely complementable singleton elements.  
This property of the partial assignment domain is required to learn 
elements that help to guide the model search away from the
conflicting region of the search space.  Complementation operator 
in abstract domains is different from the notion of precise complements. 
%
Most numerical abstract domains, such as intervals and octagons lack
complements, but they can be represented as intersections of
half-spaces, each of which has a complement. 
%
% We identify specific properties of domain elements necessary for 
% abstract model search and learning in abstractions.  
We formalise this in the sequel.
%
\begin{definition} 
A \emph{meet irreducible} $m$ in a complete lattice 
structure $A$ is an element with the following property.
\begin{equation}
\forall m_1, m_2 \in A: m_1 \meet m_2 = m \implies (m = m_1 \lor m = m_2), m \neq \top  
\end{equation}
Meet irreducibles in ACDCL correspond to the concept of literals in a SAT solver.
\end{definition}
%
The meet irreducibles in the Boolean domain $\booldomain$ for a variable $x$ are 
$x$ and $\neg x$. The meet 
irreducibles in the template polyhedra domain are all elements 
that concretise to half-spaces, i.e., they can be represented 
by a single inequality. For the interval domain, these are 
\ $x \leq d$ or $x \geq d$ for constants $d$.

@@

\begin{definition} 
An element $\absval\in \domain$ is called \emph{precisely complementable}
iff $\exists \bar{\absval} \in \domain: \neg\gamma(\bar{\absval})=\gamma(\absval)$, 
i.e., there is an element whose negated concretisation equals
the concretisation of $\absval$.
%m\meet\bar{m}=\bot \wedge m\join\bar{m}=\top$.
%A complementable meet irreducible $\bar{M}$ of an abstract lattice $A$ is the complement of a meet 
%irreducible $M \in A$ such that $\bar{M} \in A$ and the concretisation of $M$ 
%is the set complement of $\bar{M}$.  
%If every meet irreducible in $A$ is complementable, then $A$ is said to
%be have complementable meet irreducibles.
\end{definition}
